\chapter{Polynomial Evaluation}

\begin{problem}
  Given: (1) a polynomial of degree $n$ with real coefficients,
  \begin{equation*}
    p(z) = p_{n}z^{n} + \dots + p_{1}z + p_{0}
  \end{equation*}
  where $p_{n}\neq0$, and $p_{j}\in\RR$ for $j=0,\dots,n$.

  (2) A point $x_{0}$.

  Produce: the value of the polynomial at that point, $p(x_{0})$.
\end{problem}

\section{Naive Polynomial Evaluation}

The first attempt at a solution is to just ``do the obvious thing'':
loop over the coefficients, multiply them by $x^{k}$, then sum them all
up. This is summarized in Algorithm~\ref{alg:naive-polynomial-evaluation}.
The first question we should ask:
\emph{How many operations does this algorithm use?}

\begin{algorithm}\label{alg:naive-polynomial-evaluation}
  \caption{Naive polynomial evaluation.}
  \begin{algorithmic}[1]
    \Function{NaiveEval}{$p$, $x_{0}$} \Comment{$p$ is an array}
      \State\label{alg-step:naive-polynomial-evaluation:initial-load} $r\gets p_{0}$
      \State $n\gets\Call{PolynomialOrder}{p}$
      \State $x\gets 1$
      \For{$j=1,n$}
        \State $x\gets x\otimes x_{0}$
        \State\label{alg-step:naive-polynomial-evaluation:fma} $r\gets r\oplus p_{j}\otimes x$
      \EndFor
      \State \Return $r$  
    \EndFunction
\end{algorithmic}
\end{algorithm}

Each iteration of the loop has 1 addition operation, and 2
multiplication operations. This gives us a total of $n$ addition
operations and $2n$ multiplication operations.

If the computer has a fused multiplication--addition operation, then
step~\ref{alg-step:naive-polynomial-evaluation:fma} may be done in one
operation. This reduces the number of operations to $n$ additions, and
$n$ FMA operations.

\begin{rmk}[Memory loading operations]
We could also note, when we write $p_{j}$, we really mean, ``Load the
contents of memory cell associated with $p_{j}$ into a CPU register.''
If we wanted to track this as well, we would note there is an initial
load operation in
step~\ref{alg-step:naive-polynomial-evaluation:initial-load}, and in
each iteration (in step~\ref{alg-step:naive-polynomial-evaluation:fma})
there is a load operation. Hence we have $n+1$ load operations. These
are not necessarily negligible!

On x86 architectures, {\tt gfortran}\footnote{Using GNU Fortran (Ubuntu
9.3.0-17ubuntu1-20.04) 9.3.0; the flags supplied are ``\texttt{-c -S
  -fverbose-asm -O0 -c}''. On x86, the additional flags
``\texttt{-march=native -masm=intel}'' are given.} without optimization,
when compiled for Haswell architecture, it translates $p_{j}$ lookup to
a \texttt{MOV R,R} instruction, a \texttt{CDQE} instruction,
a \texttt{MOV R,R} instruction, and a \texttt{VMOVSD} instruction.
The \texttt{VMOVSD} costs 6
cycles\footnote{\url{https://uops.info/html-instr/VMOVSD_XMM_M64.html}},
the \texttt{MOV R,R} costs 1 cycle each, and the \texttt{CDQE} costs 1
cycle. Hence together they cost 9 cycles, comparable to 3 \texttt{FADD} operations.

When cross-compiling for \textsc{Sparc64}, looking up $p_{j}$
compiles\footnote{If we had a \texttt{get\_elt(p,j) = p[j]} function, this
would require an additional {\tt ldx} operation. The
difference would cost us not just 3 cycles, but also the overhead of a
function call. The point being: on \textsc{Sparc64}, accessing an array
seems to be comparable to a floating-point addition operation in CPU cycles.}
to one {\tt ldd} operation, an {\tt ld} operation, an {\tt ldx}
operation, an {\tt sra} operation, and a {\tt sllx} operation. Appendix
B3 for the UltraSPARC T1 supplement\footnote{\url{https://www.oracle.com/technetwork/systems/opensparc/t1-09-ust1-uasuppl-draft-hp-ext-1537737.html}}
gives the following latencies for these instructions: {\tt ld} and {\tt ldd}
take 9 cycles, {\tt ldx} takes 3 cycles, {\tt sra} takes 1 cycle, and
{\tt sllx} takes 1 cycle. This sums to 23 cycles, comparable to a
floating-point addition 26 cycle latency.
% fmuld - 29, faddd - 26, fdivd 83, 

An \textsc{Arm} Cortex-A9 (with hard-float support) is more explicit
taking $p_{j}$ as two \texttt{LDR} operations, followed by an
\texttt{LSLS}, an integer \texttt{ADD} of $p$ and $j$, then a
\texttt{VLDR} operation to load that memory cell as a double-precision
floating-point. These instructions may be tallied as follows:\footnote{\url{https://developer.arm.com/documentation/ddi0388/i/cycle-timings-and-interlock-behavior/load-and-store-instructions}}
3 cycles for \texttt{LDR}, 1 cycle for \texttt{ADD}, 1 cycle for \texttt{VLDR},
1 cycle for \texttt{LSLS}. Altogether, accessing the array element costs
9 cycles. This is identical to the total time it takes for \texttt{VADD}
to add two floating-point numbers and write it back to the register
file.

For x64, Haswell architecture, using Intel syntax, it compiles to:
\begin{Verbatim}
    mov     eax, DWORD PTR -20[rbp]    # tmp112, j
    cdqe
    mov     rdx, QWORD PTR -72[rbp]    # tmp113, p
    vmovsd xmm0, QWORD PTR [rdx+rax*8] # _7, *p_20(D)
\end{Verbatim}
For \textsc{Sparc64}:
\begin{Verbatim}
    ld   [%fp+2043], %g1 ! j, tmp143
    sra  %g1, 0, %g1     ! tmp143, _6
    ldx  [%fp+2175], %g2 ! p, tmp144
    sllx %g1, 3, %g1     ! _6,, tmp145
    ldd  [%g2+%g1], %f10 ! *p_20(D), _7
\end{Verbatim}
For ARM:
\begin{Verbatim}
    ldr     r2, [r7, #12]    @ tmp140, p
    ldr     r3, [r7, #52]    @ tmp141, j
    lsls    r3, r3, #3       @ tmp142, tmp141,
    add     r3, r3, r2       @ tmp143, tmp140
    vldr.64 d6, [r3]         @ _5, *p_18(D)
\end{Verbatim}
RISC-V compiles it to:
\begin{Verbatim}
    lw    a5,-20(s0)  # _6, j
    ld    a4,-72(s0)  # tmp106, p
    slli  a5,a5,3     #, tmp107, _6
    add   a5,a4,a5    # tmp107, tmp108, tmp106
    fld   fa4,0(a5)   # _7, *p_20(D)
\end{Verbatim}
MIPS compiles it to:
\begin{Verbatim}
    lw    $3,64($fp)   # tmp225, p
    lw    $2,8($fp)    # tmp226, j
    sll   $2,$2,3      # tmp227, tmp226,
    addu  $2,$3,$2     # tmp228, tmp225, tmp227
    ldc1  $f2,0($2)    # _5, *p_18(D)
\end{Verbatim}
For MIPS I-7200\footnote{Ch. 15 of \url{https://s3-eu-west-1.amazonaws.com/downloads-mips/I7200/I7200+product+launch/MIPS_I7200_Programmers_Guide_01_20_MD01232.pdf}},
these are 2 load operations (\texttt{lw}), 2 arithmetic operations
(\texttt{sll} and \texttt{addu}), and an ``\verb#ldc1#'' operation. Load
operations cost between 2 and 3 cycles, arithmetic operations cost
between 1 and 2 cycles. Altogether, this costs between 8 to 13 cycles,
or between 2 and 3 floating-point addition operations\footnote{See,
e.g., \url{https://lauri.võsandi.com/tub/computer-architecture/mips-pipeline.html}}.

These three examples show looking up $p_{j}$ can cost anywhere between 1
and 3 floating-point addition operations.
\end{rmk}