\section{Numerical Differentiation}

There are various strategies for approximating the derivative of a
function $f(x)$ at some point $x=a$. Let's just review some of the
clever ideas:
\begin{enumerate}
\item Take some polynomial interpolation $P(x)$ of $f(x)$, then use
  $P'(x)$ as the approximation for $f'(x)$. (Amazingly enough, this works.)
\item Take finitely many $f(a)$, $f(a\pm h)$, \dots, $f(a\pm kh)$, then
  take their linear combination
  $c_{-k}f(a-kh)+\dots+c_{-1}f(a-h)+c_{0}f(a)+c_{1}f(a+h)+\dots+c_{k}f(a+kh)=f'(a)+Ch^{2k}f^{(2k)}(\xi)$
  which requires solving $2k+1$ equations in $2k+1$ unknowns (the
  coefficients $c_{j}$)
\item Some clever methods using complex analysis, which is beyond our scope.
\end{enumerate}
Some caveats must be stressed, especially:
\emph{numerical differentiation is numerical unstable}. This is because
we're subtracting nearly equal values $f(x+h)-f(x)$ for $|h|\ll 1$.

\begin{defn}
  Let $f\in C^{2}[a,a+h]$.
  The \define{Forward Difference} approximation to the derivative $f'(a)$
  is given by
  \begin{equation}
    \frac{f(a+h)-f(a)}{h} = f'(a) + \frac{1}{2}f''(\xi)h
  \end{equation}
  for some $\xi\in[a,a+h]$.
\end{defn}
\begin{proof}
  Taylor expand to second order
  \begin{equation}
    f(a+h) = f(a) + hf'(a) + \frac{1}{2!}h^{2}f''(\xi)
  \end{equation}
  where $a\leq\xi\leq a+h$. Rearranging terms gives the result.
\end{proof}

\begin{defn}
  Let $f\in C^{2}[a-h,a]$.
  The \define{Backward Difference} approximation to the derivative $f'(a)$
  is given by
  \begin{equation}
    \frac{f(a)-f(a-h)}{h} = f'(a) - \frac{1}{2}f''(\xi)h
  \end{equation}
  for some $\xi\in[a-h,a]$.
\end{defn}
\begin{proof}
  Taylor expand to second order
  \begin{equation}
    f(a-h) = f(a) - hf'(a) + \frac{1}{2!}h^{2}f''(\xi)
  \end{equation}
  where $a-h\leq\xi\leq a$. We see
  \begin{equation}
    f(a-h)-f(a) = -hf'(a) + \frac{1}{2!}h^{2}f''(\xi),
  \end{equation}
  and dividing through by $-h$ gives the result.
\end{proof}

\begin{defn}
  Let $f\in C^{3}[a-h,a+h]$.
  The \define{Central Difference} approximation to the derivative $f'(a)$
  is given by
  \begin{equation}
    \frac{f(a+h)-f(a-h)}{2h} = f'(a) - \frac{1}{3}f'''(\xi)h^{2}
  \end{equation}
  for some $\xi\in[a-h,a+h]$.
\end{defn}
\begin{proof}
  Take
  \begin{subequations}
  \begin{align}
    f(a+h) &= f(a) + hf'(a) + \frac{1}{2!}h^{2}f''(a) + \frac{1}{3!}f^{(3)}(\xi_{1})h^{3}\\
    f(a-h) &= f(a) - hf'(a) + \frac{1}{2!}h^{2}f''(a)
    - \frac{1}{3!}f^{(3)}(\xi_{2})h^{3}
  \end{align}
  \end{subequations}
  for $a-h\leq\xi_{2}\leq a\leq\xi_{1}\leq a+h$.
  Then subtraction gives us
  \begin{equation}
    f(a+h)-f(a-h) = 2hf'(a) + \frac{1}{3}h^{3}\bigl(f^{(3)}(\xi_{1})+f^{(3)}(\xi_{2})\bigr).
  \end{equation}
  Applying the intermediate value theorem gives the error term
  \begin{equation}
    R(a,h)h = \frac{1}{3}h^{3}f^{(3)}(\xi)
  \end{equation}
  for some $\xi_{2}\leq \xi\leq\xi_{1}$. Rearranging terms gives the result.
\end{proof}

\begin{ex}
  Consider $f(x)=\sin(x)$. What is its approximate derivative using
  these finite-difference schemes?

  We compute, using the angle-addition formula
  \begin{equation}
    \sin(a\pm h) = \sin(a)\cos(h)+\sin(\pm h)\cos(a)
  \end{equation}
  that
  \begin{equation}
    \frac{\sin(a+h)-\sin(a)}{h} = \frac{\sin(a)\bigl(\cos(h)-1\bigr)}{h}
    + \frac{\sin(h)}{h}\cos(a).
  \end{equation}
  Similarly, for the backward difference, we have
  \begin{equation}
    \frac{\sin(a)-\sin(a-h)}{h} = \frac{\sin(a)\bigl(1-\cos(h)\bigr)}{h}
    + \frac{\sin(h)}{h}\cos(a).
  \end{equation}
  Central difference gives us
  \begin{equation}
    \begin{split}
      \sin(a+h)-\sin(a-h) &= [\sin(a)\cos(h) +\sin(h)\cos(a)]\\
      &\qquad-[\sin(a)\cos(h)-\sin(h)\cos(a)]\\
    &= 2\sin(h)\cos(a).
    \end{split}
  \end{equation}
  Hence we obtain
  \begin{equation}
    \frac{\sin(a+h)-\sin(a-h)}{2h} = \frac{\sin(h)}{h}\cos(a).
  \end{equation}
  For what values of $h$ would this be good to machine precision? (In
  other words, when would these be exact for double-precision
  arithmetic? When will round-off error be our friend, rather than arch-enemy.)
\end{ex}

\begin{xca}
  Find the forward difference, backward difference, and central
  difference approximation to the derivative of $f(x)=\exp(x)$ at some
  point $x=a$ with step-size $h$. Compare it to the exact value $f'(a)=\exp(a)$.
\end{xca}

\begin{xca}
  Find the forward difference, backward difference, and central
  difference approximation to the derivative of $f(x)=\log(x)$ at some
  point $x=a$ with step-size $h$. Compare it to the exact value $f'(a)=1/a$.
\end{xca}


\begin{xca}
  Find the forward difference, backward difference, and central
  difference approximation to the derivative of $f(x)=\tan(x)$ at some
  point $x=a$ with step-size $h$. Compare it to the exact value $f'(a)=\sec^{2}(a)$.
\end{xca}

\begin{xca}
  How do these finite-difference schemes respect the usual laws of
  differentiation? That is, if we have $h_{1} = g\circ f$, what's the error
  of the chain-rule? What about the quotient rule $h_{2}(x)=g(x)/f(x)$?
\end{xca}

\begin{puzzle}
Is there a ``best'' step size $h$ for numerical differentiation? (In the
sense of minimizing error.)
\end{puzzle}

The basic algorithm for answering this question is the same in each
approximation for numerical differentiation. We'll work out the solution
for the central difference scheme.

\begin{thm}
  The optimal step-size for central difference approximation of $f'(a)$ is
  \begin{equation}
    h = \left|\frac{3\machEps |f(a)|}{|f'''(\xi)|}\right|^{1/3},
  \end{equation}
  where $\xi\in[a-h,a+h]$.
\end{thm}
\begin{proof}
  We want to find
  \begin{equation}
    E(h) = \left|f'(a) - \frac{\fl(f(a+h))-\fl(f(a-h))}{2h}\right|.
  \end{equation}
  Let
  \begin{equation}
    \frac{\fl(f(a+h) - f(a-h))}{2h} = \frac{f(a+h)-f(a-h)}{2h} + \frac{\delta(a+h)-\delta(a-h)}{2h}
  \end{equation}
  be the round-off error, due to using floating-point arithmetic. Note
  that $\delta(a)\approx f(a)\machEps$ is on the order of $\ulp(f(a))$.
  Then
  \begin{equation}
     E(h) = \left|f'(a) - \frac{f(a+h)-f(a-h)}{2h} - \frac{\delta(a+h)-\delta(a-h)}{2h}\right|
  \end{equation}
  is the error. This is bounded by
  \begin{equation}
     E(h) \leq\left|f'(a) - \frac{f(a+h)-f(a-h)}{2h}\right|+ \left|\frac{\delta(a+h)-\delta(a-h)}{2h}\right|
  \end{equation}
  using the triangle inequality, and then
  \begin{equation}
    \left|f'(a) - \frac{f(a+h)-f(a-h)}{2h}\right|+ \left|\frac{\delta(a+h)-\delta(a-h)}{2h}\right|
    \leq\frac{h^{2}M}{6} + \frac{2\machEps}{2h}|f(a)|
  \end{equation}
  using the error for central difference, and round-off error being
  bounded by machine-epsilon by the fundamental axiom of floating-point arithmetic.
  Thus have have our desired bound
  \begin{equation}\label{eq:quadrature:derivatives:error-bound}
    E(h) \leq\frac{h^{2}M}{6} + \frac{2\machEps}{2h}|f(a)|.
  \end{equation}
  Since this is an upper bound, we can just pretend they're
  equal\footnote{This is common in mathematical analysis: work with
  bounds, then minimize the upper bound, and sandwich estimates together.}.
  We find the extrema of this error by taking the derivative
  \begin{equation}
    E'(h_{\text{cr}})=\frac{h_{\text{cr}}M}{3}-\frac{\machEps |f(a)|}{h_{\text{cr}}^{2}}=0.
  \end{equation}
  Solving this gives us
  \begin{equation*}
    h_{\text{cr}} = \left|\frac{3\machEps |f(a)|}{M}\right|^{1/3}.\qedhere
  \end{equation*}
\end{proof}

\begin{xca}
  Determine a good estimate for $f'''(\xi)$ in the previous theorem. Can
  we use finite differences? Polynomial interpolation?
\end{xca}
\begin{xca}
  Prove, for the forward difference approximation of $f'(a)$, that the
  optimal step size is $h = \sqrt{4\machEps |f(a)|/|f''(\xi)|}$.
\end{xca}
\begin{xca}
  From the bounds on error given in Eq~\eqref{eq:quadrature:derivatives:error-bound},
  determine the stability of central difference.
\end{xca}